---
title: "JHU4"
author: "bhaskar"
output: html_document
---

*Exploratory graphs*
1. univariate
2. bi variate
3. multivariate

Exploratory statistics
1. Five number summary
2. variance
3. covariance correlation
4. quantiles

Exploratory graphing system
1. base
```{r}
library(datasets)
head(airquality)
hist(airquality$Ozone)
boxplot((airquality$Ozone))
with(airquality,plot(Wind,Ozone))
airquality = transform(airquality,Month=as.factor(Month))
boxplot(Ozone~Month,airquality,xlab="Month",ylab="Ozone")
```




**important params** *pch,lwd,lty,col,xlab,ylab*
**important par() params** *mar,mfow,mfcol,oma,bg,las*
*plot,lines,points,title,mtext,text,axis*

### **base**

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", 
    type = "n"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
with(subset(airquality, Month != 5), points(Wind, Ozone, col = "red"))
legend("topright", pch = 1, col = c("blue", "red"), legend = c("May", "Other Months"))
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", 
    pch = 20))
model <- lm(Ozone ~ Wind, airquality)
abline(model, lwd = 2)
```


**multiple plots**
```{r}
par(mfrow = c(1, 2))
with(airquality, {
    plot(Wind, Ozone, main = "Ozone and Wind")
    plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
})
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0))
with(airquality, {
    plot(Wind, Ozone, main = "Ozone and Wind")
    plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
    plot(Temp, Ozone, main = "Ozone and Temperature")
    mtext("Ozone and Weather in New York City", outer = TRUE)
})
```


2. Lattice
```{r}
library(lattice)
xyplot(Ozone ~ Wind | Month, data = airquality, layout = c(5, 1))

```

3. ggplot2

## qplot

```{r}
library(ggplot2)
str(mpg)
```

**scatter plots**

```{r}
qplot(displ, hwy, data = mpg)
qplot(displ, hwy, data = mpg, color = drv)
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"))
```

** histogram **
```{r}
qplot(hwy, data = mpg, fill = drv)
```

**facets**
```{r}
qplot(displ, hwy, data = mpg, facets = . ~ drv)
qplot(hwy, data = mpg, facets = drv ~ ., binwidth = 2)
```

## GGPLOT2 GRAMMER
- A data frame
- aesthetic mappings: how data are mapped to color, size
- geoms: geometric objects like points, lines, shapes.
- facets: for conditional plots.
- stats: statistical transformations like binning, quantiles, smoothing.
- scales: what scale an aesthetic map uses (example: male = red, female = blue).
- coordinate system

## Clustering

```{r}
set.seed(1234)
par(mar = c(0, 0, 0, 0))
x <- rnorm(12, mean = rep(1:3, each = 4), sd = 0.2)
y <- rnorm(12, mean = rep(c(1, 2, 1), each = 4), sd = 0.2)
plot(x, y, col = "blue", pch = 19, cex = 2)
text(x + 0.05, y + 0.05, labels = as.character(1:12))

```

** hclust **

** Distance or similarity for both hclust and kmeans**
- Continuous - euclidean distance
- Continous - correlation similarity
- Binary - manhattan distance

Pick a distance/similarity that makes sense for your problem

```{r}
dataFrame <- data.frame(x = x, y = y)
distxy <- dist(dataFrame,method = "euclidian")
hClustering <- hclust(distxy)
plot(hClustering)
```

Gives an idea of the relationships between variables/observations
The picture may be unstable
- Change a few points
- Have different missing values
- Pick a different distance
- Change the merging strategy
- Change the scale of points for one variable

But it is deterministic, Choosing where to cut isn't always obvious, Should be primarily used for exploration 

## K-means

```{r}
set.seed(1234); par(mar=c(0,0,0,0))
x <- rnorm(12,mean=rep(1:3,each=4),sd=0.2)
y <- rnorm(12,mean=rep(c(1,2,1),each=4),sd=0.2)
plot(x,y,col="blue",pch=19,cex=2)
text(x+0.05,y+0.05,labels=as.character(1:12))
```

kmeans

```{r}
dataFrame <- data.frame(x,y)
kmeansObj <- kmeans(dataFrame,centers=3)
names(kmeansObj)

```

**plot**
```{r}
par(mar=rep(0.2,4))
plot(x,y,col=kmeansObj$cluster,pch=19,cex=2)
points(kmeansObj$centers,col=1:3,pch=3,cex=3,lwd=3)
```

### Heatmaps

```{r}
set.seed(1234)
dataMatrix <- as.matrix(dataFrame)[sample(1:12),]
kmeansObj <- kmeans(dataMatrix,centers=3)
par(mfrow=c(1,2), mar = c(2, 4, 0.1, 0.1))
image(t(dataMatrix)[,nrow(dataMatrix):1],yaxt="n")
image(t(dataMatrix)[,order(kmeansObj$cluster)],yaxt="n")

```

# CASE STUDY

# CLUSTERING

```{r}
load("~/mywork/JHU/data/samsungData.rda")
names(samsungData)[1:12]
head(samsungData$activity)
str(samsungData$activity)
table(samsungData$activity)
```

** plotting average acceleration for first subject**

```{r}
par(mfrow = c(1, 2), mar = c(5, 4, 1, 1))
samsungData <- transform(samsungData, activity = factor(activity))
sub1 <- subset(samsungData, subject == 1)
plot(sub1[, 1], col = sub1$activity, ylab = names(sub1)[1])
plot(sub1[, 2], col = sub1$activity, ylab = names(sub1)[2])
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), 
    pch = 1)
```

## hclust

** clustering based on average acceleration**
```{r}
par(mfrow = c(1,1))
myplclust <- function(hclust, lab = hclust$labels, lab.col = rep(1, length(hclust$labels)), 
    hang = 0.1, ...) {
    ## modifiction of plclust for plotting hclust objects *in colour*!  Copyright
    ## Eva KF Chan 2009 Arguments: hclust: hclust object lab: a character vector
    ## of labels of the leaves of the tree lab.col: colour for the labels;
    ## NA=default device foreground colour hang: as in hclust & plclust Side
    ## effect: A display of hierarchical cluster with coloured leaf labels.
    y <- rep(hclust$height, 2)
    x <- as.numeric(hclust$merge)
    y <- y[which(x < 0)]
    x <- x[which(x < 0)]
    x <- abs(x)
    y <- y[order(x)]
    x <- x[order(x)]
    plot(hclust, labels = FALSE, hang = hang, ...)
    text(x = x, y = y[hclust$order] - (max(hclust$height) * hang), labels = lab[hclust$order], 
        col = lab.col[hclust$order], srt = 90, adj = c(1, 0.5), xpd = NA, ...)
}
distanceMatrix <- dist(sub1[, 1:3])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col = unclass(sub1$activity))
```

*not much of seperation with various activities, which is deduced from mix of all the colors in the bottom of graph*

** plotting max acceleration **

```{r}
par(mfrow = c(1, 2))
plot(sub1[, 10], pch = 19, col = sub1$activity, ylab = names(sub1)[10])
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), 
    pch = 20)
plot(sub1[, 11], pch = 19, col = sub1$activity, ylab = names(sub1)[11])
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), 
    pch = 20)
par(mfrow = c(1,1))
```

** clustering based on max acceleleration **

```{r}
distanceMatrix <- dist(sub1[, 10:12])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col = unclass(sub1$activity))
```

*seperated out walking activities and unmoving activities*

** singular value decomposition **

```{r}
svd1 = svd(scale(sub1[, -c(562, 563)]))
par(mfrow = c(1, 2))
plot(svd1$u[, 1], col = sub1$activity, pch = 19)
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), 
    pch = 20)
plot(svd1$u[, 2], col = sub1$activity, pch = 19)
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), 
    pch = 20)
par(mfrow = c(1, 1))
```

** finding maximum contributor**

```{r}
plot(svd1$v[, 2], pch = 19)
maxContrib <- which.max(svd1$v[, 2])
```

** clustering based on maximum contributor **

```{r}

distanceMatrix <- dist(sub1[, c(10:12, maxContrib)])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col = unclass(sub1$activity))
```

** maximum contributor **

```{r}
names(samsungData)[maxContrib]
```

*adding the maximun contributor to clustering helped in additonal seperation between walking activities,still the non moving activities are not seperated*

## Kmeans

** nstart=1 first try **
```{r}
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6)
table(kClust$cluster, sub1$activity)
```

** nstart =1 second try **

```{r}
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 1)
table(kClust$cluster, sub1$activity)
```

** nstart 100 1st try **

```{r}
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 100)
table(kClust$cluster, sub1$activity)
```

** nstart 100 second try **

```{r}
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 100)
table(kClust$cluster, sub1$activity)
```

** cluster1 variable centering **

```{r}
plot(kClust$center[1, 1:10], pch = 19, ylab = "Cluster Center", xlab = "")
```

*** cluster2 variable centering **

```{r}
plot(kClust$center[4, 1:10], pch = 19, ylab = "Cluster Center", xlab = "")
```

* cluster centers gives an idea of what features are useful to find that cluster*

# Air pollution

## Has fine particle pollution in the U.S. decreased from 1999 to 2012?

## Read in data from 1999
```{r}
pm0 <- read.table("~/mywork/JHU/data/RD_501_88101_1999-0.txt", comment.char = "#", header = FALSE, sep = "|", na.strings = "")
dim(pm0)
head(pm0)
cnames <- readLines("~/mywork/JHU/data/RD_501_88101_1999-0.txt", 1)
print(cnames)
cnames <- strsplit(cnames, "|", fixed = TRUE)
print(cnames)
names(pm0) <- make.names(cnames[[1]])
head(pm0)
x0 <- pm0$Sample.Value
class(x0)
str(x0)
summary(x0)
mean(is.na(x0))
```

### Are missing values important here?

## Read in data from 2012

```{r}
pm1 <- read.table("~/mywork/JHU/data/RD_501_88101_2012-0.txt", comment.char = "#", header = FALSE, sep = "|", na.strings = "", nrow = 1304290)
names(pm1) <- make.names(cnames[[1]])
head(pm1)
dim(pm1)
x1 <- pm1$Sample.Value
class(x1)
```

### Five number summaries for both periods

```{r}
summary(x1)
summary(x0)
mean(is.na(x1))
```

### Are missing values important here?

### Make a boxplot of both 1999 and 2012

```{r}
boxplot(x0, x1)
boxplot(log10(x0), log10(x1))
```
* SO OVERALL PARTICULAR MATTER DECREASED FROM 1999 to 2012 , but there are negative values*


### Check negative values in 'x1'

```{r}
summary(x1)
negative <- x1 < 0
sum(negative, na.rm = T)
mean(negative, na.rm = T)
dates <- pm1$Date
str(dates)
dates <- as.Date(as.character(dates), "%Y%m%d")
str(dates)
hist(dates, "month")
```

*negative values in all first six monts of 2012, but percentage is less so ignoring*

### Check what's going on in months 1--6


### Plot a subset for one monitor at both times

### Find a monitor for New York State that exists in both datasets
```{r}
site0 <- unique(subset(pm0, State.Code == 36, c(County.Code, Site.ID)))
site1 <- unique(subset(pm1, State.Code == 36, c(County.Code, Site.ID)))
site0 <- paste(site0[,1], site0[,2], sep = ".")
site1 <- paste(site1[,1], site1[,2], sep = ".")
str(site0)
str(site1)
both <- intersect(site0, site1)
print(both)
```

### Find how many observations available at each monitor

```{r}
pm0$county.site <- with(pm0, paste(County.Code, Site.ID, sep = "."))
pm1$county.site <- with(pm1, paste(County.Code, Site.ID, sep = "."))
cnt0 <- subset(pm0, State.Code == 36 & county.site %in% both)
cnt1 <- subset(pm1, State.Code == 36 & county.site %in% both)
sapply(split(cnt0, cnt0$county.site), nrow)
sapply(split(cnt1, cnt1$county.site), nrow)
```

### Choose county 63 and side ID 2008

```{r}
pm1sub <- subset(pm1, State.Code == 36 & County.Code == 63 & Site.ID == 2008)
pm0sub <- subset(pm0, State.Code == 36 & County.Code == 63 & Site.ID == 2008)
dim(pm1sub)
dim(pm0sub)
```

### Plot data for 2012

```{r}
dates1 <- pm1sub$Date
x1sub <- pm1sub$Sample.Value
plot(dates1, x1sub)
dates1 <- as.Date(as.character(dates1), "%Y%m%d")
str(dates1)
plot(dates1, x1sub)
```

### Plot data for 1999

```{r}
dates0 <- pm0sub$Date
dates0 <- as.Date(as.character(dates0), "%Y%m%d")
x0sub <- pm0sub$Sample.Value
plot(dates0, x0sub)
```

### Plot data for both years in same panel

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(dates0, x0sub, pch = 20)
abline(h = median(x0sub, na.rm = T))
plot(dates1, x1sub, pch = 20)  
abline(h = median(x1sub, na.rm = T))
```

*adjust for diferent range for 1999 and 2012*

### Find global range

```{r}
rng <- range(x0sub, x1sub, na.rm = T)
rng
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(dates0, x0sub, pch = 20, ylim = rng)
abline(h = median(x0sub, na.rm = T))
plot(dates1, x1sub, pch = 20, ylim = rng)
abline(h = median(x1sub, na.rm = T))
```

### Show state-wide means and make a plot showing trend

```{r}
head(pm0)
mn0 <- with(pm0, tapply(Sample.Value, State.Code, mean, na.rm = T))
str(mn0)
summary(mn0)
mn1 <- with(pm1, tapply(Sample.Value, State.Code, mean, na.rm = T))
str(mn1)
names(mn0)
names(mn1)
```

### Make separate data frames for states / years

```{r}
d0 <- data.frame(state = names(mn0), mean = mn0)
d1 <- data.frame(state = names(mn1), mean = mn1)
mrg <- merge(d0, d1, by = "state")
dim(mrg)
head(mrg)
```

## Connect lines

```{r}
par(mfrow = c(1, 1))
with(mrg, plot(rep(1, 52), mrg[, 2], xlim = c(.5, 2.5)))
with(mrg, points(rep(2, 52), mrg[, 3]))
segments(rep(1, 52), mrg[, 2], rep(2, 52), mrg[, 3])
```


